\vspace{-1cm}\normalsize{\textbf{Fairly Random: Impact of Winning the Toss on the Probability of Winning}}

### 2015-12-16 write_up/cricket.tex
updated
update on rank stuff --- not happy yet
more writing

In nearly all cricket matches, it is claimed that there is a clear advantage to either bowling or batting first. The advantage is pointed to by commentators, by the team captains in the pre-toss interview, and by the captain of the losing team in the post-match interview. And to wrest the said advantage, one merely needs to pick the side of the coin left facing the sky after the toss. And while this method of granting advantage is fair on average, the system isn't fair in any one game.

At first glance, the imbalance seems inevitable. After all, someone has to bat first. One can, however, devise a baseball like system in which short innings are interspersed. If that violates the nature of the game too much, one can easily create pitches that don't deteriorate appreciably over the course of a game. Or, one can come up with an estimate of the advantage, and adjust the scores accordingly, akin to an adjustment issued when matches are shortened due to rain.

@@

To assess the net observed advantage of winning the toss, we exploit data from a novel dataset of over 43,000 first-class men's cricket matches --- to our knowledge, the largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}. And in analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into. To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat, Saad2015} etc., we do not condition on post coin-toss decisions. We find that winning the toss grants a small but significant advantage, but that advantage varies considerably and systematically.

@@

We assume that the outcome of a toss is random. Conditional on the outcome of a toss being random, the effect of winning the toss can be attributed to the toss itself. Before we exploit the design, we shed some light on the validity of the assumption. In particular, we assess whether the coin toss is somehow rigged, with the home side enjoying the rub of the green more often. For this analysis, we only get to exploit international matches as establishing which of the teams is the home team in local matches is somewhat arduous. Of the 5,591 international matches, the home team won the toss in 2,860 matches, or about 51.15\%. The probability that we will get as large a proportion by chance after tossing 5,591 coins is about 8.7\%, low but not eyebrow-raisingly so. Thus, for now, we choose to trust that the tosses are random.

@@

The team that wins the toss wins the match 2.9\% more often than the team that loses the toss. This is a reasonable sized advantage in a competitive sport --- though likely much smaller than the number that most commentators carry in their heads. This advantage, however, varies by format, by conditions, by whether or not a particular formula was used to adjust scores when it rains, and how much better the team that won the toss is vis-\`{a}-vis the competing team. Much of the variability follows expected patterns.

@@

\begin{figure}[htbp]
\centering
\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Difference in Ranks}
\includegraphics[scale=.85]{../figs/winbyRank.pdf}
\label{fig:ranks}
\end{figure}

>>>>

In nearly all cricket matches, it is claimed that there is a clear advantage to either bowling or batting first. The advantage is pointed to by commentators, by the team captains in the pre-toss interview, and by the captain of the losing team in the post-match interview. And to wrest the said advantage, the captain merely needs to pick the side of the coin that will be left facing the sky after the toss. And while this method of granting advantage is fair on average, the system isn't fair in any one game.

At first glance, this imbalance seems inevitable. After all, someone has to bat first. One can, however, devise a baseball like system, with a series of short innings woven together. If that violates the nature of the game too much, one can easily create pitches that don't deteriorate appreciably over the course of a match. Or, one can come up with an estimate of the advantage, and adjust the scores accordingly, akin to an adjustment that is issued when the matches are shortened due to rain.

@@

To assess the net observed advantage of winning the toss, we exploit data from a novel dataset of over 43,000 first-class men's cricket matches --- to our knowledge, the largest ever dataset assembled for the question, and nearly 50--100 times larger than used in prominent previous attempts \citep[see,][]{dawson2009bat, de1998winning}. In analyzing these data, we avoid a common but important pitfall that some other studies on the topic fall into. To avoid post-treatment bias \citep[see][]{acharya2015}, unlike \citet{dawson2009bat}, \citet{Saad2015}, etc., we do not condition on post coin-toss decisions. We find that winning the toss grants a small but significant advantage, but that advantage varies considerably and systematically.

@@

We assume that the outcome of a toss is random. Conditional on the outcome of a toss being random, the effect of winning the toss can be attributed to the toss itself. Any decision made after the outcome of the toss is known, however, is `post-treatment.' In particular, the decision to bat or bowl first is made after accounting for the relative strengths and weaknesses vis-\`{a}-vis the competing team at that particular instance, and thus not independent of team attributes. Hence, conditioning on decision to bowl or bat first can induce bias in estimates of advantage of winning a toss. Thus, unlike \citet{dawson2009bat}, \citet{Saad2015}, we solely rely on the assumption that the outcome of a toss is random.

But before we exploit the design, we shed some light on the validity of the assumption. In particular, we assess whether the coin toss is somehow rigged, with the home side enjoying the rub of the green more often. For this analysis, we only get to exploit international matches as establishing which of the teams is the home team in local matches is somewhat arduous. Of the 5,591 international matches, the home team won the toss in 2,860 matches, or about 51.15\%. The probability that we will get as large a proportion by chance after tossing 5,591 coins is about 8.7\%, low but not eyebrow-raisingly so. Thus, for now, we choose to trust that the tosses are random.

@@

The team that wins the toss wins the match 2.9\% more often than the team that loses the toss. This is a reasonable sized advantage in a competitive sport --- though likely much smaller than the number that most commentators carry in their heads.
% Also interpret it by comparing to probability of winning when playing with the worst placed team (not defined from last team)
This advantage, however, varies by format, by conditions, by whether or not a particular formula was used to adjust scores when it rains, and how much better the team that won the toss is vis-\`{a}-vis the competing team. Much of the variability follows expected patterns.

@@

\begin{figure}[htbp]
\centering
\caption{Percentage of Matches Won Minus Matches Lost After Winning the Toss by Difference in Ranks}
\includegraphics[width=1\textwidth]{../figs/winbyRank.pdf}
\label{fig:ranks}
\end{figure}
